Forcing for Isabelle

### 2022-10-24 Independence_CH/conclusion.tex
Some comments inspired by Tim Chow added

\section*{Acknowledgments}
\label{sec:acknowledgments}
We thank Z.~Vidnyánszky for some discussions where the idea of
building the summary in \theory{Definitions\_Main} took shape.
Mikhail Mandrykin helped us with technical assistance regarding the
rendering of formulas in the Isabelle/jEdit interface, which improved
our presentation of this material during the Latin American Congress
of Mathematicians (CLAM 2021); we extend our gratitude to the rest of
the Isabelle community, for their kind support through the Users mailing
list. We also want to warmly thank Larry Paulson for his work and
encouragement.  Last, and definitely not least, we are deeply grateful
to Ken Kunen for his
inspiring expositions. This formalization is a tiny homage to his
memory.

>>>>

\section*{Acknowledgments}
\label{sec:acknowledgments}
We are indebted to Timothy Chow for insightful comments on a draft of
this paper. We also thank Z.~Vidnyánszky for some discussions where the idea of
building the summary in \theory{Definitions\_Main} took shape.
Mikhail Mandrykin helped us with technical assistance regarding the
rendering of formulas in the Isabelle/jEdit interface, which improved
our presentation of this material during the Latin American Congress
of Mathematicians (CLAM 2021); we extend our gratitude to the rest of
the Isabelle community, for their kind support through the Users mailing
list. We also want to warmly thank Larry Paulson for his work and
encouragement.  Last, and definitely not least, we are deeply grateful
to Ken Kunen for his
inspiring expositions. This formalization is a tiny homage to his
memory.

### 2022-10-24 Independence_CH/related.tex
Minor edits
Some comments inspired by Tim Chow added

@@

This faithfulness to set-theoretical practice does not come for
free. Recursive constructions and inductive definitions are far easier
to perform in the Calculus of Inductive Constructions on which Lean
is based, and in Isabelle/ZF are rather cumbersome. Also, a
typed discipline provides aid to write succinctly and many assumptions
are satisfied by mere notation. To be clear, those benefits come from
doing set theory in a non set-theoretical meta-language. On the other
hand, Isar proofs, as the one shown in
Sec.~\ref{sec:sample-formal-proof}, are easier to write and understand
than the language of tactics of Lean.

@@

>>>>

This is perhaps an appropriate time to insist that we have \emph{not}
formalized the relative consistency of $\neg\CH$, and we are actually
not aiming for that in the short term. In our context, going for the
plain consistency result seems off the mark, since we can not weaken
our base theory (which is essentially equivalent to $\ZF$)%%,
%% and we catch our tail.
Even if we intended to do so, the standard route to use ctms for a
relative consistency proof (through the Reflection Theorem) is
prohibitive for us, since our metatheory does not have the required
induction principles (on Isabelle/ZF formulas---in contrast to
formulas coded as sets, as in our presentation).

@@

This faithfulness to set-theoretical practice does not come for
free. Recursive constructions and inductive definitions are far easier
to perform in the Calculus of Inductive Constructions on which Lean
is based, and in Isabelle/ZF are rather cumbersome. Also, a
typed discipline provides aid to write succinctly and many assumptions
are satisfied by mere notation. To be clear, those benefits come from
doing set theory in a non set-theoretical language. On the other
hand, Isar proofs, as the one shown in
Sec.~\ref{sec:sample-formal-proof}, are easier to write and understand
than the language of tactics of Lean.

@@

Concerning the minimum amount of Replacement needed to construct
forcing extensions, only recently we learned about Mathias' work on the
subject (for which a summary is offered in
\cite[Sect.~6]{kanamori-mathias}). In
\cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set Theory
are offered for which each of the inclusions $M\subseteq M[G]$ and
$M\supseteq M[G]$ fail, where the poset $\pP$ is the trivial $\{\1
\}$. In one of them, $K$, we have $\om\in K \sm K[G]$, hence the
ordinals do not coincide.

Also, in the reference \cite{mathias-provident}, a reasonably minimal
fragment \textsc{Prov} of $\ZF$ that allows to do set forcing is
identified, and transitive sets satisfying it are called
\emph{provident}. Existence of rank and of transitive closure are
among them; hence their appearance in our list seems
justified. Nevertheless, \textsc{Prov} is far weaker than the
fragments of $\ZF$ considered here, since it is a restriction of
Kripke-Platek Set Theory, and thus it does not include neither
Powerset nor full Separation. The detailed theory of provident sets is
developed in \cite{mathias-bowler-gentle}. % Its Section~0 is very
% readable.

### 2022-10-24 notes/chow.tex
Some comments inspired by Tim Chow added

It is perhaps important to make clear that we have not obtained the
relative consistency of $\neg\CH$, and we are
actually not aiming for that in the short term. In our context, going
for the plain consistency relative to $\ZF$ seems off the mark, since we
can not weaken our base theory (which is essentially equivalent to $\ZF$)%% ,
%% and we catch our tail.
Even if we intended to do so, the standard route to use ctms for a
relative consistency proof (through the Reflection Theorem) is
prohibitive for us, since our metatheory does not have the required
induction principles (on Isabelle/ZF formulas---in contrast to
formulas coded as sets, as in our presentation).

@@

In \cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set
Theory are offered that each of the inclusions $M\subseteq M[G]$ and
$M\supseteq M[G]$ fail, where the poset $\pP$ is the trivial
$\{\1 \}$. In one of them, $K$, we have $\om\in K \sm K[G]$, hence the
ordinals do not coincide.

>>>>

It is perhaps important to make clear that we have not formalized the
relative consistency of $\neg\CH$, and we are
actually not aiming for that in the short term. In our context, going
for the plain consistency result seems off the mark, since we
can not weaken our base theory (which is essentially equivalent to $\ZF$)%% ,
%% and we catch our tail.
Even if we intended to do so, the standard route to use ctms for a
relative consistency proof (through the Reflection Theorem) is
prohibitive for us, since our metatheory does not have the required
induction principles (on Isabelle/ZF formulas---in contrast to
formulas coded as sets, as in our presentation).

@@

In \cite[Sect.~1]{mathias:hal-01188043}, models $M$ of Zermelo Set
Theory are offered for which each of the inclusions $M\subseteq M[G]$ and
$M\supseteq M[G]$ fail, where the poset $\pP$ is the trivial
$\{\1 \}$. In one of them, $K$, we have $\om\in K \sm K[G]$, hence the
ordinals do not coincide.
